---
title: "HW03-P1 Adv Map Viz"
format: 
  html: 
    embed-resources: true
---


## Introduction

The following exercises will use census tract `tidycensus` data for Ramsey and Hennepin county and crash data (2019-2022) from the [pedestrian/bike crash database](https://information.stpaul.gov/datasets/stpaul::pedestrian-and-bike-crash-dataset-/about) for St. Paul within Ramsey county in the Twin Cities.

**Requirements:**

- Explore ACS and enter an addition variable to code
- Download and Read in roads data
- Ensure both census2020 and crashes are same CRS
- Crop roads (at least to Ramsey and Hennepin counties)
- Join crashes and census to calculate number of crashes per tract
- Join census2020 and crashes per tract
- Choropleth map of number of crashes per tract + use of scale_fill_gradientn()
- Choropleth map coloring census tracts by a census variable + crash locations as points + scale_fill_gradientn()
- Sentence Interpretation in Data Context
- 2 non-map visuals of number of crashes vs census variables
- Sentence Interpretation in Data Context

## 0 Request U.S. Census Data API Key

In order to retrieve data from the U.S. Census website, and API key is required.  Follow the steps below to obtain yours and use it in this homework:

- Go to <https://api.census.gov/data/key_signup.html> and request an API key.
- Save the received key into a file along side this file with the name `census_api_key.txt`--make sure that the file contains nothing other than the key.
- Open GitHub Desktop -> right click `census_api_key.txt` and select Ignore file (add to .gitignore) to avoid exposing your key to others

## 1 Load One More Variable

The following code loads census data for Ramsey counties and reads in the pedestrian/bike crash data.  Adjust this code to load in one additional census variable that you think might serve as useful context for crash data. To do this:

- Look at the documentation for `get_acs()` by entering `?tidycensus::get_acs` in the Console.
- Look at the [Searching for variables documentation](https://walker-data.com/tidycensus/articles/basic-usage.html#searching-for-variables) from the `tidycensus` package. Use the function in this documentation to find one variable to add to the `variables` argument in the `get_acs()` function below.
- Rename the variable that you picked to have a more descriptive name within `rename()`.

> Enter the new variable you selected (e.g., B01003_001) here: B01002_001
> 
> Enter the new name for that variable (e.g., population) here: medianAge

```{r}
library(tidyverse)
library(sf)
```

```{r reading}
#| message: false
#| warning: false
#| output: false


census2020 <- tidycensus::get_acs(
    year = 2020,
    state = "MN",
    geography = "tract",
    variables = c("B01003_001", "B01002_001", "B19013_001", "B23006_023", "B25058_001", "B25107_001", "B25003_001", "B25003_002", "B25003_003", "B25077_001"),
    output = "wide",
    geometry = TRUE,
    key = readLines("census_api_key.txt")

) |>
    filter(word(NAME, 4) %in% c("Ramsey")) |>
    mutate(
        tract = word(NAME, 3),
        tract = str_remove(tract, ","),
        county = word(NAME, 4)
    ) |>
    select(-NAME) |>
    rename(
        "population" = "B01003_001E",
        "medianAge" = "B01002_001E",
        "medianIncome" = "B19013_001E",
        "bachelors" = "B23006_023E",
        "medContractRent" = "B25058_001E",
        "tenureTotal" = "B25003_001E",
        "tenureOwned" = "B25003_002E",
        "tenureRented" = "B25003_003E",
        "medianHomeValue"= "B25077_001E"
    ) |>
    select(-contains("_"))

crashes <- read_csv("https://mac-stat.github.io/data/Crash.csv") |>
    filter(!is.na(Latitude), !is.na(Longitude))
```

## 2 Download Shapefile

Visit [DATA.GOV](https://catalog.data.gov/dataset/tiger-line-shapefile-2019-state-minnesota-primary-and-secondary-roads-state-based-shapefil?from_hint=eyJxIjoibWlubmVzb3RhLXByaW1hcnktYW5kLXNlY29uZGFyeS1yb2FkcyIsInNvcnQiOiJyZWxldmFuY2UifQ%3D%3D) and download the Shapefile Zip File (`tl_2019_27_prisecroads.zip`) to the `data/raw` folder of this homework then unzip.

Load in the roads shapefile using `st_read()`, and transform `roads` to have the same CRS as `census2020` if necessary.

```{r reading_roads}
roads <- st_read("data/raw/tl_2019_27_prisecroads/tl_2019_27_prisecroads.shp", quiet = TRUE)

# Check CRS of roads and transform if necessary
roads <- st_transform(roads, st_crs(census2020))

```


## 3 Transofrm to `sf` Object

Transform the `crashes` data frame to a `sf` object that has point geometry using the code below.

```{r crashes_to_sf}
crashes <- st_as_sf(crashes, coords = c("Longitude", "Latitude"), crs = "NAD83")
```


## 4 Check CRSs

Check if the CRSs are the same for `census2020` and `crashes` using `st_crs()`. Transform to ensure they are exactly the same.

```{r same_crs}
st_crs(census2020) == st_crs(crashes)
crashes <- st_transform(crashes, st_crs(census2020))
st_crs(census2020) == st_crs(crashes)
```

## 5 Crop

Use `st_crop()` to crop the roads and census2020 datasets to the area we are interested in (crashes in St. Paul within Ramsey County). To determine the region to crop to, you will need to use `st_bbox()` on a spatial data object that only pertains to St. Paul within Ramsey County.

```{r crop_roads}
crash_area <- st_bbox(crashes)
roads_sub <- st_crop(roads, crash_area)
census2020_sub <- st_crop(census2020, crash_area)
```

## 6 Join Spatial Datasets

Join the crashes and census dataset together and count the number of crashes per census tract. The function `st_join` can join to spatial data sets according to whether the spatial geometries of the right table intersect with the spatial geometries of the left table. It matters which order you join the datasets in!

```{r join_crashes}
crashes_per_tract <- st_join(crashes, census2020_sub |> select(GEOID)) |>
    st_drop_geometry() |> # removes geometry - makes the following calculation more efficient 
    filter(!is.na(Accident_Datetime)) |>
    count(GEOID) 
```

## 7 Join Census

Join the census data with `crashes_per_tract` and then use a filter of `n > 0` to only keep the census tracts where crashes were recorded instead of all of Ramsey County. Again, order matters!

```{r add_geometry_back}
crashes_per_tract_geo <- census2020_sub |> # sf object with census geometry goes first
    left_join(crashes_per_tract, by = "GEOID") |>
    filter(n > 0)
```

## 8 Pedestrian/Bike Crashes per Census Tract

Create a map that shows the number of pedestrian/bike crashes per census tract. Write one sentence about the patterns you notices; interpret your findings in the data context.

- Use `scale_fill_gradientn()` to set up a meaningful color scale for the census tract colors.
- Include a map layer for the roads.

```{r map_num_crashes}
ggplot() +
    geom_sf(data = crashes_per_tract_geo, aes(fill = n), color = "white") +
    geom_sf(data = roads_sub, color = "gray40", linewidth = 0.2) +
    scale_fill_gradientn(colors = c("white", "orange", "red")) +
    labs(
        title = "Pedestrian and Bike Crashes in St. Paul",
        fill = "Crashes"
    ) +
    theme_void()
```

The tracts with the most recorded pedestrian and bike crashes are concentrated in a smaller number of parts of St. Paul rather than being evenly distributed across the city.


## 9 Crash Locations

Create one map that shows crash locations as points and colors census tracts by one different census variable. Write one sentence about the patterns you notice; interpret your findings in the data context.

- Use `scale_fill_gradientn()` to set up a color scale for the census tract colors.
- Include a layer for the roads.

```{r mas_crashes_census_info}
ggplot() +
    geom_sf(data = census2020_sub, aes(fill = medianIncome), color = "white") +
    geom_sf(data = roads_sub, color = "gray50", linewidth = 0.2) +
    geom_sf(data = crashes, color = "black", size = 0.7, alpha = 0.6) +
    scale_fill_gradientn(colors = c("lightyellow", "gold", "darkgreen")) +
    labs(
        title = "Crash Locations and Median Household Income",
        fill = "Median income") +
    theme_void()
```

Crash locations appear in tracts with a range of median household incomes, so the map does not suggest that income alone explains where crashes occur.

## 10 Number of Crashes and Census Data

Create two other visuals (that are not maps) showing the relationships between the number of crashes and census data. Write one sentence about the patterns you notice; interpret your findings in the data context.

```{r crashes_census_info}
ggplot(crashes_per_tract_geo, aes(x = medianIncome, y = n)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    labs(
        title = "Crashes and Median Household Income",
        x = "Median household income",
        y = "Number of crashes")

ggplot(crashes_per_tract_geo, aes(x = medianAge, y = n)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    labs(
        title = "Crashes and Median Age",
        x = "Median age",
        y = "Number of crashes")
```

The points are spread out around both fitted lines, suggesting that neither median income nor median age has a strong simple linear relationship with the number of recorded crashes in a tract.


## Resources Reflection (required)

List the resources you used to help with this assignment then write 3-5 sentence reflection on which resources were most helpful in finishing this task.

**Resources:**

- Course notes on spatial data and advanced maps
- `tidycensus::get_acs()` documentation and variable search
- `sf` documentation for `st_transform()`, `st_crop()`, and `st_join()`

**Reflection:**

The course spatial-data notes were the most helpful resource because they showed how CRS values and `sf` objects work together. The `tidycensus` documentation helped me find median age and identify its ACS variable code. The `sf` documentation helped me decide which dataset should come first in the spatial join. This assignment helped me understand that a map usually requires several preparation steps before the plotting code.
