---
title: "7 Adv Data wrangling II"
---



## 🧩 Learning Goals

By the end of this lesson, you should be able to:

- Manipulate and explore strings using the `stringr` package
- Construct regular expressions to find patterns in strings


## Motivation

In 2018 the data journalism organization [The Pudding](https://pudding.cool/) featured a story called [30 Years of American Anxieties](https://pudding.cool/2018/11/dearabby/) highlighting themes in 30 years of posts to the Dear Abby column (an American advice column).

To understand themes in text data, one can conduct a **qualitative text analysis** which is a manual process that involved multiple people reading through the instances of text several time until they reach a consensus about themes.  Given the amount of labor involved, this process is suitable for medium datasets.

For large datasets, one might want to use **quantitative & computational text analysis** which leverages computers, data science techniques, and natural language process to extract themes from the data.  We will explore this approach in this lesson.

Both qualitative and computational approaches can be used in tandem. Often, using computational tools can help focus a close reading of select texts, which parallels the spirit of a qualitative analysis.

To prepare ourselves for a computational analysis, let's learn about strings.


## Strings

Strings are objects of the `character` class (abbreviated as `<chr>` in tibbles).

When you print out strings, they display with double quotes:

```{r chr_class}
some_string <- "banana"
some_string
```

Working with strings generally will involve the use of **regular expressions**, a tool for finding patterns in strings.

Regular expressions (regex, for short) look like the following:

```
"^the" <-- Find strings that start with "the"
"end$" <-- Find strings that end with "end"
```

Before getting to regular expressions, let's go over some fundamentals about working with strings. The `stringr` package (which a package in the `tidyverse` ecosystem) is great for working with strings.

```{r doc_setup_07}
#| message: false

library(tidyverse) # The stringr package is part of tidyverse
```


### Create Strings

Creating strings by hand is useful for testing out regular expressions.

To create a string, type any text in either double quotes `"double"` or single quotes `'single'`. Using double or single quotes doesn't matter unless your string itself has single or double quotes.

```{r create_strings}
string1 <- "This is a string"
string2 <- 'If I want to include a "quote" inside a string, I use single quotes'
string3 <- c(string1, string2) # string / character vector (of greater than length 1)


class(string1)
class(string2)
class(string3)

length(string1)
length(string2)
length(string3)
```

We can view these strings "naturally" (without the opening and closing quotes) with `str_view()`:

```{r str_view}
str_view(string1)
str_view(string2)
str_view(string3)
```


### ✏️ Quote String

Create the string `It's Thursday`. What happens if you put the string inside single quotes? Double quotes?

```{r ex_create_string}
quote_string <- "It's Thursday"
quote_string

```


### Escape Characters

Because `"` and `'` are special characters in the creation of strings, R offers another way to put them inside a string. We can **escape** these special characters by putting a `\` in front of them:

```{r escape_chars}
string1 <- "This is a string with \"double quotes\""
string2 <- "This is a string with \'single quotes\'"
str_view(string1)
str_view(string2)
```

Given that `\` is a special character, how can we put the `\` character in strings? We have to escape it with `\\`.


### ✏️ Escape Backslash

Create the string `C:\Users`. What happens when you don't escape the `\`?

```{r ex_create_string_slash}
path_string <- "C:\\Users"
path_string

```


### Special Characters

Other special characters include:

- `\t` <- creates a tab
- `\n` <- creates a newline

Both can be useful in plots to more neatly arrange text.

```{r escape_tab_newline}
string1 <- "Record temp:\t102"
string2 <- "Record temp:\n102"

str_view(string1)
str_view(string2)
```

Can we get `str_view()` to show the tab instead of `{\t}`? We can use the `html` argument to have the string displayed as if on a webpage:

```{r str_view_html}
str_view(string1, html = TRUE)
```


### Combine

Often we will want to create new strings within data frames. We can use `str_c()` or `str_glue()`, both of which are vectorized functions, ie, they take vectors as inputs and provide vectors as outputs which means that they can be used within `mutate()`

- With `str_c()` the strings to be combined are all arguments separated by commas.
- With `str_glue()` the desired string is written as a template with variable names inside curly braces `{}`.

```{r str_c_glue}
df <- tibble(
    first_name = c("Arya", "Olenna", "Tyrion", "Melisandre"),
    last_name = c("Stark", "Tyrell", "Lannister", NA)
)
df
```

```{r}
df |>
    mutate(
        full_name1 = str_c(first_name, " ", last_name),
        full_name2 = str_glue("{first_name} {last_name}")
    )
```

### ✏️ Build Date Strings

In the following data frame, create a full date string in month-day-year format using both `str_c()` and `str_glue()`.

```{r ex_str_c_glue}
df_dates <- tibble(
    year = 2000:2002,
    month = c("Jan", "Feb", "Mar"),
    day = 3:5
)
df_dates
```

```{r}
df_dates |>
  mutate(
    date1 = str_c(month, "-", day, "-", year),
    date2 = str_glue("{month}-{day}-{year}")
  )

```


### Length

The `str_length()` counts the number of characters in a string.

```{r str_length}
comments <- tibble(
    name = c("Alice", "Bob"),
    comment = c("The essay was well organized around the core message and had good transitions.", "Good job!")
)
comments
```

```{r}
comments |>
    mutate(
        comment_length = str_length(comment)
    )
```

### Substring

The `str_sub()` function gets a substring of a string. The 2nd and 3rd arguments indicate the beginning and ending position to extract.

- Negative positions indicate the position from the end of the word, eg, -3 indicates "3rd letter from the end"
- Specifying a position that goes beyond the word won't result in an error. `str_sub()` will just go as far as possible.

```{r str_sub}
x <- c("Apple", "Banana", "Pear")
x
```

```{r}
str_sub(x, start = 1, end = 3)
str_sub(x, start = -3, end = -1)
str_sub(x, start = 2, end = -1)
str_sub("a", start = 1, end = 15)
```

### ✏️ Find Middle Letter

Using `str_sub()`, create a new variable with only the middle letter of each word in the data frame below. (Challenge: How would you handle words with an even number of letters?)

```{r ex_str_sub}
df <- tibble(
    word_id = 1:3,
    word = c("replace", "match", "pattern")
)
df
```

```{r}
df |>
  mutate(
    middle_position = ceiling(str_length(word) / 2),
    middle_letter = str_sub(word, middle_position, middle_position)
  )

```


## Patterns

Suppose that you're exploring text data looking for places where people describe happiness. There are many ways to search. We could search for the word "happy" but that excludes "happiness" so we might search for "happi".

**Regular expressions** (regex) are a powerful language for describing patterns within strings.

```{r load_regex_data}
# load few datasets
data(fruit)
sample(fruit, 10)
```

```{r}
data(words)
sample(words, 10)
```

```{r}
data(sentences)
sample(sentences, 10)
```

We can use `str_view()` with the `pattern` argument to see what parts of a string match the regex supplied in the `pattern` argument. (Matches are enclosed in `<>`.)

```{r str_view_pattern}
str_view(fruit, "berry")
```


### Quantifiers

- Letters and numbers in a regex are matched exactly and are called **literal characters**.
- Most punctuation characters, like `.`, `+`, `*`, `[`, `]`, and `?`, have special meanings and are called **metacharacters**.
- **Quantifiers** are metacharacters that specify how many times the proceeding character (group or character class) must occurs for a match to be found:
    - `?`: the proceeding character must occurs 0 or 1 times for a match to be found
    - `+`: the proceeding character must occurs 1+ (1 or more times) for a match to be found
    - `*`: the proceeding character must occurs 0+ (0 or more times) for a match to be found


### ✏️ Predict Quantifier Matches

Before running the code below, predict what matches will be made. Run the code to check your guesses. Note that in all regex's below the `?`, `+`, `*` applies to the `b` only (not the `a`).

```{r ex_regex_quantifiers}
str_view(c("a", "ab", "abb"), "ab?")
```

```{r}
str_view(c("a", "ab", "abb"), "ab+")
```

```{r}
str_view(c("a", "ab", "abb"), "ab*")
```

### Character Classes

- We can match any of a set of characters with `[]` (called a **character class**), e.g., `[abcd]` matches "a", "b", "c", or "d".
- We can invert the match by starting with `^`: `[^abcd]` matches anything except "a", "b", "c", or "d".

```{r regex_classes}
# Match words that have vowel-x-vowel
str_view(words, "[aeiou]x[aeiou]")

# Match words that have not_vowel-y-not_vowel
str_view(words, "[^aeiou]y[^aeiou]")
```


### ✏️ Find Vowel Pairs

Using the `words` dataset, find words that have two vowels in a row followed by an "m".

```{r ex_regex_classes}
str_view(words, "[aeiou][aeiou]m")
```


### Alternation

- The **alternation** operator `|` can be read just like the logical operator `|` ("OR") to pick between one or more alternative patterns, eg, `apple|banana` searches for "apple" or "banana".

```{r regex_alternation}
str_view(fruit, "apple|melon|nut")
```


### ✏️ Find Repeated Vowels

Using the `fruit` dataset, find fruits that have a repeated vowel ("aa", "ee", "ii", "oo", or "uu".)

```{r ex_regex_alternation}
str_view(fruit, "aa|ee|ii|oo|uu")

```


### Anchors & Groups

- The `^` operator indicates the beginning of a string, and the `$` operator indicates the end of a string, eg, `^a` matches strings that start with "a", and `a$` matches words that end with "a".
- Parentheses group together parts of a regular expression that should be taken as a bundle. Much like parentheses in arithmetic statements, eg, `ab+` is a little confusing. Does it match "ab" one or more times? Or does it match "a" first, then just "b" one or more times? (The latter, as we saw in an earlier example.) We can be very explicit and use `a(b)+`.


### ✏️ Find Words with y

Using the `words` dataset, find words that start with "y"

```{r ex_regex_start_end_group}
str_view(words, "^y")

```

Using the `words` dataset again, find words that don't start with "y".

```{r ex_regex_start_end_group_2}
str_view(words, "^[^y]")

```


## Dear Abby Letters

### Load the Letters

Read in the "Dear Abby" data underlying The Pudding's [30 Years of American Anxieties](https://pudding.cool/2018/11/dearabby/) article.

```{r read_dear_abby}
posts <- read_csv("https://raw.githubusercontent.com/the-pudding/data/master/dearabby/raw_da_qs.csv")
str(posts)
```

Take a couple minutes to scroll through the [30 Years of American Anxieties](https://pudding.cool/2018/11/dearabby/) article to get ideas for themes that you might want to search for using regular expressions.

### Regex Functions

The following are core `stringr` functions that use regular expressions:

- `str_view()` - View the first occurrence in a string that matches the regex
- `str_count()` - Count the number of times a regex matches within a string
- `str_detect()` - Determine if (TRUE/FALSE) the regex is found within string
- `str_subset()` - Return subset of strings that match the regex
- `str_extract(), str_extract_all()` - Return portion of each string that matches the regex. `str_extract()` extracts the first instance of the match. `str_extract_all()` extracts all matches.
- `str_replace(), str_replace_all()` - Replace portion of string that matches the regex with something else. `str_replace()` replaces the first instance of the match. `str_replace_all()` replaces all instances of the match.
- `str_remove(), str_remove_all()` - Removes the portion of the string that matches the pattern. Equivalent to `str_replace(x, "THE REGEX PATTERN", "")`


### ✏️ Search for Themes

Pick a theme from the article and use each of the functions above, starting with `str_count()`, to search for it in the `posts` data. Pull up each function's documentation page and read through the arguments as you go.

```{r}
theme <- "marriage|married|husband|wife"

head(str_count(posts$question_only, theme))
head(str_detect(posts$question_only, theme))
head(str_subset(posts$question_only, theme))
head(str_extract(posts$question_only, theme))
head(str_extract_all(posts$question_only, theme))
head(str_replace(posts$question_only, theme, "relationship"))
head(str_replace_all(posts$question_only, theme, "relationship"))
head(str_remove(posts$question_only, theme))
head(str_remove_all(posts$question_only, theme))
```


## Done!

- Check the ICA Instructions for how to (a) push your code to GitHub and (b) update your portfolio website
