---
title: "9 Functions"
---



## 🧩 Learning Goals

By the end of this lesson, you should be able to:

- Recognize when it would be useful to write a function
- Identify the core components of a function definition and explain their role (the `function()` directive, arguments, argument defaults, function body, return value)
- Identify the scoping of a function in accessing objects in R
- Describe the difference between argument matching by position and by name
- Write `if`-`else`, `if`-`else if`-`else` statements to conditionally execute code
- Write your own function to carry out a repeated task
- Provide feedback on functions written by others



## Functions

### Why Functions?

Getting really good at writing useful and reusable functions is one of the best ways to increase your expertise in data science. It requires a lot of practice.

If you've copied and pasted code **3 or more times**, it's time to write a function. Avoid repeating ourselves results in many benefits such as:

1. **Reducing errors:** Copy-and-paste then modify is prone to errors, eg, forgetting to change a variable name.
2. **Efficiency:** If you need to update code, you only need to do it in one place. This allows reuse of code within and across projects.
3. **Readability:** Encapsulating code within a function with a descriptive name makes code more readable.


### Core Parts

When you define a function, what does it look like?

```r
# Defining a function
function_name <- function(input_name_1, input_name_2 = default_value_2){

  # function body - code that does something

  return(output)
}

# Calling a function in all valid ways
function_name(input_name_1 = 2, input_name_2 = 4)
function_name(2, 4)
function_name(input_name_2 = 4, input_name_1 = 2)
function_name(2)
```

The core parts of defining a function include:

- **`function()` directive**
    - This is what tells R to create a function.
- **Arguments/Inputs**: the `input_name_1` and `input_name_2` -- these are names of the function inputs
    - `default_value_2` is a default value that is used if no input values are provided when the function is called
- **Function body**
    - The code inside the curly braces `{` `}` is where all the work happens. This code uses the function arguments to perform computations.
- **Return value**
    - We can explicitly return an object by putting it inside `return()`.


### Properties

Here are three examples that can help us learn a few properties of functions:

```{r}
average <- function(x, remove_nas) {
    return(sum(x, na.rm = remove_nas)/length(x))
}

average2 <- function(x, remove_nas = TRUE) {
    return(sum(x, na.rm = remove_nas)/length(x))
}

average3 <- function(x, remove_nas) {
    sum(x, na.rm = remove_nas)/length(x)
}
```

Note:

- In `average2`, the `remove_nas` argument has a **default** value of `TRUE`.

When a function has default values for arguments, they don't have to be provided when you call the function if you want to use the default value. See below:

```{r}
#| error: true

# Does throw an error
average(c(1, 2, 3, NA))
average(c(1, 2, 3, NA), remove_nas = TRUE)

# Doesn't throw an error
average2(c(1, 2, 3, NA))
average2(c(1, 2, 3, NA), remove_nas = FALSE)
```

- If we don't provide a `return()`, the last value that gets evaluated in the **function body** and isn't stored as an object is what the function returns. (This is generally the last line without an assignment operator `<-`.)
    - `average3()` is one example, but this can easily lead to errors.
    - We should explicitly return an object by putting it inside `return()`.


## Scope

In programming, **scope** refers to the area of a program where a named object is recognized and can be used.  R uses **lexical scoping** which means that the scope of a variable is determined by where it is created in the code.

Lexical scoping in R works as follows:

- When code refers to an object, eg, data set, function, and vector, R looks for the object in the current **local environment**.
- If it doesn't find it, it continues to search by looking in the parent environment, and so on. The top-level environment is the **global environment**, ie, the location where all interactive computation and storage (outside of a function) takes place. If you save an object in R outside a function local environment, it is stored in the global environment.
- If R can't find it in the global environment, it will look for it in loaded packages.

When writing and using functions, the local environment is within the function itself.  If you define an object in the function, it won't be accessible outside that function unless you pass it as the output as shown in the example below:

```{r}
#| error: true
#| output: asis

average_new <- function(x, remove_nas = TRUE) {
  sum(x, na.rm = remove_nas)/length(x)
  fun_new_thing_within_function <- "Fun times!"
}

average_new(1:3)
fun_new_thing_within_function
```

Scoping is important to consider if you try to refer to objects that aren't passed as arguments.  This can be **dangerous** because if you re-use names of variables, it is easy to accidentally refer to a variable that is not the one you intended as shown in example below:

```{r}
#| error: true
#| output: asis

v <- c(1,2,3)

average_new2 <- function(x, remove_nas = TRUE) {
  sum(x, na.rm = remove_nas)/length(x) + v
}


average_new2(1:3)
v <- c(4,5,6)
average_new2(1:3)
```

Anything created/saved/updated within a function that you want accessible outside that function needs to be passed to `return()` as shown in example below:

```{r}
#| error: true
#| output: asis

v <- c(1,2,3)

average_new3 <- function(x, remove_nas = TRUE) {
  v <- v + 10 # this is not changing v outside the function...
  return(sum(x, na.rm = remove_nas)/length(x))
}

average_new3(1:3)
v
```

For more about lexical scoping in R, see [R Programming for Data Science](https://bookdown.org/rdpeng/rprogdatascience/scoping-rules-of-r.html)


## Write Functions

Tips for writing functions:

- Write the code for the body of the function (the task that you are wanting to repeat)
- Identify the parts of the function that could/would change (these are the inputs)
    - Update the body of the function to be in terms of the inputs
    - Avoid manually typing anything that is specific to one input; use code to get that (such as unique values of a variable)
- Identify the output you want to return


### ✏️ Rescale

Write a function that rescales a numeric vector to be between 0 and 1. Below are some test cases with the expected output; use them to test your function.

- `x = 2:4`. Expected output: `0.0 0.5 1.0`
- `x = c(-1, 0, 5)`. Expected output: `0.0000000 0.1666667 1.0000000`
- `x = -3:-1`. Expected output: `0.0 0.5 1.0`

```{r}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  return((x - rng[1]) / (rng[2] - rng[1]))
}

rescale01(2:4)
rescale01(c(-1, 0, 5))
rescale01(-3:-1)
```


### ✏️ Format Phone Numbers

Write a function that formats a 10-digit phone number nicely as `(###) ###-####`. Your function should work on the following test cases: `c("651-330-8661", "6516966000", "800 867 5309")`. It may help to refer to the `stringr` cheatsheet.

```{r}
format_phone <- function(phone) {
  numbers <- stringr::str_remove_all(phone, "[^0-9]")
  return(stringr::str_replace(numbers, "(\\d{3})(\\d{3})(\\d{4})", "(\\1) \\2-\\3"))
}

format_phone(c("651-330-8661", "6516966000", "800 867 5309"))
```


## Call Functions

```{r doc_setup_09}
#| message: false

library(tidyverse)
```

When you supply arguments to a function, they can be matched **by position** and/or **by name**.

When you call a function without `argument = value` inside the parentheses, you are using **positional matching** as in the example below:


```{r}
ggplot(diamonds, aes(x = carat, y = price)) + 
    geom_point()
```

The above code works because the first argument of `ggplot` is `data` and the second is `mapping`. (Pull up the documentation on `ggplot` with `?ggplot` in the Console.) However, the following code doesn't work:

```{r}
#| error: true

ggplot(aes(x = carat, y = price), diamonds) + 
    geom_point()
```

However, if we name the arguments, we would be using the **name matching** approach and the code would work as shown below:

```{r}
ggplot(mapping = aes(x = carat, y = price), data = diamonds) + 
    geom_point()
```

Somewhat confusingly, we can name some arguments and not others. Below, `mapping` is named, but `data` isn't. This works because when an argument is matched by name, it is "removed" from the argument list, and the remaining unnamed arguments are matched in the order that they are listed in the function definition. **Just because this is possible doesn't mean it's a good idea--please, don't do this!**

```{r}
ggplot(mapping = aes(x = carat, y = price), diamonds) + geom_point()
```

::: {.callout-tip title="Argument matching"}
In general, it is safest to match arguments by name and position for your peace of mind. For functions that you are very familiar with (and know the argument order), it's ok to just use positional matching.
:::


### ✏️ Diagnose Error

Diagnose the error message in the example below:

```
ggplot() |>
    geom_sf(census_data, aes(fill = population))

Error in `layer_sf()`:
! `mapping` must be created by `aes()`
```

`geom_sf()` treats its first unnamed argument as `mapping`, so it incorrectly
tries to use `census_data` as an aesthetic mapping. Name the data argument:

```r
ggplot() +
  geom_sf(data = census_data, aes(fill = population))
```


## The `if`-`else if`-`else` Control Structure

Often in functions, you will want to execute code chunks conditionally. In a programming language, **control structures** are parts of the language that allow you to control what code is executed. By far the most common is the `if`-`else if`-`else` structure.

```r
if (logical_condition) {
    # some code
} else if (other_logical_condition) {
    # some code
} else {
    # some code
}

middle <- function(x) {
    mean_x <- mean(x, na.rm = TRUE)
    median_x <- median(x, na.rm = TRUE)
    seems_skewed <- (mean_x > 1.5*median_x) | (mean_x < (1/1.5)*median_x)
    if (seems_skewed) {
        median_x
    } else {
        mean_x
    }
}
```

::: {.callout-tip title="To vectorize or not to vectorize"}
The `if () else {}` code is not **vectorized**; the logical condition in `if()` cannot be a logical vector of length longer than 1. If you want a vectorized version of conditional execution of code use `if_else()` or `case_when()`; these can be used in `mutate()`.
:::


### ✏️ Convert Temp

Write a function for converting temperatures that takes as input a numeric value and a unit (either "C" for Celsius or "F" for Fahrenheit). The function should convert the temperature from one unit to the other based on the following formulas:

- To convert Celsius to Fahrenheit: `(Celsius * 9/5) + 32`
- To convert Fahrenheit to Celsius: `(Fahrenheit - 32) * 5/9`

```{r}
# write your answer below
```


### ✏️ Extract Domain Names

Write a function that extracts the domain name of a supplied email address. The function should return the domain name (e.g., "gmail.com"). If the input is not a valid email address, return "Invalid Email". (A valid email ends in "dot something".)

```{r}
# write your answer below
```


## Write Functions with `tidyverse` Verbs

Perhaps we are using `group_by()` and `summarize()` a lot to compute group means. We might write this function:

```{r}
group_means <- function(df, group_var, mean_var) {
    df |>
        group_by(group_var) |>
        summarize(mean = mean(mean_var))
}
```

Let's use it on the `diamonds` dataset to compute the mean size (`carat`) by diamond `cut`:

```{r}
#| error: true

group_means(diamonds, group_var = cut, mean_var = carat)
```

What is the problem? Is it that the variable names need to be in quotes?

```{r}
#| error: true

group_means(diamonds, group_var = "cut", mean_var = "carat")
```


### Tidy Evaluation

What's going on? The `tidyverse` uses something called **tidy evaluation**: this allows you to refer to a variable by typing it directly (e.g., no need to put it in quotes).

- `group_by(group_var)` is expecting a variable that is actually called `group_var`, and `mean(mean_var)` is expecting a variable that is actually called `mean_var`.

To fix this we need to **embrace** the variables inside the function with `{{` `var` `}}`:

```{r}
group_means <- function(df, group_var, mean_var) {
    df |>
    group_by({{ group_var }}) |>
    summarize(mean = mean({{ mean_var }}))
}
```

The `{{` `var` `}}` tells R to look at what the variable specified by the input  `var` rather than look for a variable called `var`.

```{r}
group_means(diamonds, group_var = cut, mean_var = carat)
```


### Pick

Let's group by both `cut` and `color`:

```{r}
#| error: true

group_means(diamonds, group_var = c(cut, color), mean_var = carat)
```

Oh No! What Now?! When `c(cut, color)` is put inside `{{` `c(cut, color)` `}}` within the function, R is actually running the code inside `{{` `}}`.

- This combines the columns for those 2 variables into one long vector. What we really meant by `c(cut, color)` is "group by both cut and color".

To fix this, we need the `pick()` function to get R to see `{{` `group_var` `}}` as a vector of separate variables (like the way `select()` works).

```{r}
group_means <- function(df, group_var, mean_var) {
    df |>
    group_by(pick({{ group_var }})) |>
    summarize(mean = mean({{ mean_var }}))
}
```


### ✏️ Count with Proportions

Create a new version of `dplyr::count()` that also shows proportions instead of just sample sizes. The function should be able to handle counting by multiple variables. Test your function with two different sets of arguments using the `diamonds` dataset.

```{r}
# write your answer below
```


### ✏️ Plot with Smoothers

Create a function that creates a scatterplot from a user-supplied dataset with user-supplied x and y variables. The plot should also show a curvy smoothing line in blue, and a linear smoothing line in red. Test your function using the `diamonds` dataset.

```{r}
# write your answer below
```


## Done!

- Check the ICA Instructions for how to (a) push your code to GitHub and (b) update your portfolio website
