---
title: "Dead Sea Scrolls Manuscripts"
author: "Xingyuan Xu"
format: html
---

## Question

Which identified works appear most frequently in the collection, and how does their canonical status differ?

## Data

The data come from the [TidyTuesday project](https://github.com/rfordatascience/tidytuesday/tree/main/data/2026/2026-09-15). The original source is the Israel Antiquities Authority's Leon Levy Dead Sea Scrolls Digital Library.

```{r}
library(tidyverse)
library(ggplot2)

dead_sea_scrolls <- read_csv("dead_sea_scrolls.csv")
```

## Visualization

```{r}
book_counts <- count(
  filter(
    dead_sea_scrolls,
    !is.na(biblical_book),
    canon_status != "Unidentified"),
  biblical_book,
  canon_status, sort = TRUE)
book_counts <- slice_head(book_counts, n = 10)

ggplot(
  book_counts,
  aes(
    x = fct_reorder(biblical_book, n),
    y = n,
    fill = canon_status)) +
  geom_col(width = 0.60) +
  geom_hline(
    yintercept = mean(book_counts$n),
    linetype = "dashed",
    color = "gray") +
  scale_fill_manual(
    values = c(
      "Protocanonical" = "darkred",
      "Deuterocanonical" = "lightyellow",
      "Non-canonical" = "lightblue3")) +
  labs(
    title = "Psalms and Deuteronomy appear most frequently",
    subtitle = "Ten most common identified works in the collection",
    x = NULL,
    y = "Number of manuscripts",
    fill = "Canonical status")+
    coord_flip() +
  theme_classic()
```

Psalms and Deuteronomy are the two most frequently represented identified works in the dataset. The chart also shows that frequently represented works include both protocanonical texts and non-canonical works, such as Phylactery.
