---
title: "Changes in Urban Green Area"
author: "Xingyuan Xu"
format: html
---

## Question

Which cities had the largest increases and decreases in the share of green area between 1990 and 2025?

## Data

The data come from the [TidyTuesday project](https://github.com/rfordatascience/tidytuesday/tree/main/data/2026/2026-09-22) and were originally published by the UN-Habitat Urban Indicators Database.
Green area includes long-term vegetation such as trees, forests, shrubs, and perennial grasses, but does not include water.

```{r}
#| message: false

library(tidyverse)

urban <- read_csv("urban_2026-09-22.csv")
```

## Visualization

```{r}
green_change <- urban %>%
  select(countryOrTerritoryName, cityName, year,
         averageShareOfGreenAreaInCityUrbanAreaPct) %>%
  filter(year %in% c(1990, 2025)) %>%
  pivot_wider(
    names_from = year,
    values_from = averageShareOfGreenAreaInCityUrbanAreaPct,
    names_prefix = "year_") %>%
  filter(!is.na(year_1990), !is.na(year_2025)) %>%
  mutate(
    change = year_2025 - year_1990,
    city = paste(cityName, countryOrTerritoryName, sep = ", "))

cities_to_plot <- bind_rows(
  slice_min(green_change, change, n = 5),
  slice_max(green_change, change, n = 5)) %>%
  mutate(direction = if_else(change > 0, "Increase", "Decrease"))

ggplot(cities_to_plot,
       aes(x = fct_reorder(city, change), y = change, fill = direction)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Decrease" = "darkorange",
                               "Increase" = "darkgreen")) +
  labs(
    title = "Bissau had the largest decline in green-area share",
    subtitle = "Five largest increases and decreases from 1990 to 2025",
    x = NULL,
    y = "Change in green area (percentage points)",
    fill = NULL,
    caption = "Source: UN-Habitat via TidyTuesday") +
  theme_classic()
```

Bissau had by far the largest decrease among cities with data in both years, falling by about 63 percentage points.
The largest increases were much smaller: Leon had the greatest increase at about 15 percentage points.
These values describe changes within the urban boundaries used by UN-Habitat and do not by themselves explain why the changes occurred.
